Experience Report: Haskell in Computational Biology

### 2012-02-08 mrfy_experience_report.tex
"beautiful code" shitty draft done

breadcrumb: HOFs for search.

@@

>>>>

Just as the Viterbi algorithm is one major component of MRFy, stochastic search 
is the other. In designing MRFy, we did not know in advance what sort of 
stochastic search technique would converge the fastest, or give us the best 
alignments with the fastest performance. We decided to implement several 
different search strategies so that we could experiment with all of them. We 
found this to be easy and pleasant thanks to Haskell's support of higher order 
functions.

We wrote a basic search function in only twenty-five lines of code; as one of 
its parameters we provide a SearchStrategy data type, which we use to define 
specific search techniques. We require a SearchStrategy to define four 
functions: mutate, accept, terminate, and initialize; we use the Haskell module 
system to allow sharing of some of these functions between different search 
strategies. For example, both \textit{random mutation hill climb} and 
\textit{simulated annealing} use the same initialize and mutate functions, but 
very different accept functions. In all, we implemented four search strategies: 
\textit{random mutation hill climb}, \textit{simulated annealing}, 
\textit{multistart simulated annealing}, and \textit{genetic algorithm}. We 
found the implementation of the search function and search strategies to be 
pleasant, and the resulting code to be simple and elegant.

In MRFy, the Viterbi function is called several times (once for each 
non-beta-strand region) for each search generation. Furthermore, in a 
population-based search such as \textit{multistart simulated annealing}, the 
scoring function (which itself calls Viterbi) is called many times per 
generation. Beyond traditional profiling, we found parallelization to be an 
obvious way to improve runtime performance. We found this to be trivial thanks 
to the Control.Parallel library. We pass the Viterbi function to a map over all 
non-beta-strand regions of the model and query sequence, and these regions are 
independent of one another. Similarly, we pass the scoring function to a map 
over the population of potential solutions, which are also independent. We 
merely substituted ``parmap rseq'' for map. MRFy now efficiently uses all six 
cores on an AMD workstation for a non-population-based simulated annealing 
search, and all forty-eight cores on a compute server for population-based 
searches.

@@

- profiling
-
